### 2015-01-08 doc/thesis/conclusions.tex
Rewriting the conclusions and adding an abstract

Our main contribution is the design of a complete optional type system
for Lua, a procedural scripting language.
TypeScript is a complete optional type system for JavaScript,
but it does not include union types, and it uses arrays to represent
variadic functions and multiple return values.
Gradualtalk is a complete gradual type system for Smalltalk,
but Smalltalk is an object-oriented language that does
not handle multiple return values, multiple assignments, and
variadic functions.
Gradualtalk also mixes nominal and structural typing because of
the object-oriented nature of Smalltalk.
Typed Racket is a complete static type system for Racket, a functional
language.
Typed Racket makes it easier to migrate from untyped to typed code,
but the migration happens module-by-module and type annotations are
mandatory on typed modules.
There are research on the design of optional and gradual type systems
for JavaScript, Perl, and Python, but they are either experimental or
limited in their coverage.
Furthermore, Typed Lua includes table types that use types as keys,
instead of identifiers, and the fact that Typed Lua also includes
literal types makes its table types subsume the usual idea of
record types.

Although each language is different, we believe that several parts of
the design of Typed Lua should be relevant to other scripting languages.
Most notably, JavaScript shares with Lua the use of tests of type tags
as a way of encoding optional and overloaded parameters, and the
proliferation of class-based object systems built on top of the
language's delegation mechanisms.

Typed Lua should also be a major contribution to the Lua community,
because it offers a framework that programmers can use to document,
test, and better structure their applications.
For libraries where a full conversion to static type checking should
prove unfeasible or too much work, the community can use Typed Lua
just to document the external interfaces of the libraries,
giving the benefits of static type checking to the users of these
libraries.

Finally, Typed Lua can be a base for further research and development
of static type systems for the Lua language, such as the use of static
analysis of Lua API calls for checking that a native library is
exporting the interface to Lua code that it declares, the use of
effect types for typing Lua coroutines, and the use of static types
for code optimization.

>>>>

We presented Typed Lua, an optional type system for Lua.
Typed Lua is a Lua extension that allows programmers to combine static
and dynamic typing in Lua code, making easier the evolution of simple scripts
into large programs.

Our main contribution is the formalization of a complete optional
type system for Lua.
Even though Lua shares several characteristics with other dynamically
typed languages such as JavaScript, Lua also has several unusual features
that are not present in the type system of these languages.
Lua has functions with flexible arity that interact with multiple
assignment, functions that are overloaded on the number of return values,
and incremental evolution of record and object types.

We introduce the refinement of table types to support the incremental
evolution of record and object types.
The refinement of table types plays an important role in our type
system, as they static type check the idiomatic way that Lua programmers
use tables to define modules and objects.

We introduce projection types to handle functions that are overloaded
on the number of return values.
Projection types are a novel feature of our type system, as they
allow programmers to narrow the type of a local variable that depends
on the type of another one.

We use union types and variadic types to allow our type system to
handle functions with flexible arity.
While union types can describe optional parameters, variadic types
describe the type of the vararg expression and the type of functions
that can receive or return any number of values.

Unlike some of the other optional type systems, we designed Typed Lua
without deliberated unsound parts.
However, we still do not have proofs that the novel parts of our type
system are sound.

To improve our confidence on the design of Typed Lua, we performed several
case studies to evaluate it.
Before starting the design of Typed Lua, we realized that we should
not rely only on the semantics of Lua, as this could lead to a
cumbersome type system that would not support several idiomatic Lua code.
For this reason, we performed a mostly automated survey of Lua idioms
and features to support our design choices.

Even though the evaluation results exposed several limitations of our
type system, they also gave us confidence that our design still
introduces a reasonable amount of static type safety to several Lua idioms
and features.
We found that the three main limitations of our type system are
the lack of intersection types, parametric polymorphism, and operator overloading.
We see a soundness proof along with these limitations as our major targets
for future work.
While a soundness proof can consolidate our confidence on our design choices,
overcoming these limitations can increase the amount of static type safety
that our type system provides.

We believe that Typed Lua is a major contribution to the Lua community,
because it offers a framework that programmers can use to document,
test, and better structure their applications.
For libraries where a full conversion to static type checking should
prove unfeasible or too much work, the community can use Typed Lua
just to document the external interfaces of the libraries,
giving the benefits of static type checking to the users of these
libraries.
In fact, we already have user feedback from Lua programmers that are
using Typed Lua in their projects.
For instance, ZeroBrane Studio is an IDE for Lua development that is
starting to use Typed Lua to perform static analysis on Lua code.

Finally, we also believe that Typed Lua can be a base for further research
and development of static type systems for the Lua language, such as
the use of static analysis of Lua API calls for checking that a native library is
exporting the interface to Lua code that it declares, the use of
effect types for typing Lua coroutines, and the use of static types
for code optimization.
